---
title: "Occupancy models"
author: "F. Barraquand"
date: "January 16, 2024"
output:
  html_document:
    highlight: textmate
    theme: united
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: '4'
---
  
```{r setup, include=FALSE}
options(width = 300)
knitr::opts_chunk$set(cache = FALSE) 
library(R2jags)
library(knitr)
library(mcmcplots)
library(ggmcmc)

```

Example taken from Kéry \& Schaub's *Bayesian Population Analysis* 2011 book. 

## Gathering the data

* 27 sites (woodpiles), 6 replicated counts for each. 

* Covariates: `forest_edge` (edge or more interior), `date`, `hour` (date and hour of day)

* Detection at 10 of 27 woodpiles and from 1 to 5 times

* Questions:

    - Have some bluebugs been likely missed in some sites? I.e., is there a discrepancy between *observed* number of sites with bluebugs and *estimated* number of sites with bluebugs. 
    - How many times should one visit a woodpile to detect bluebugs ? $\rightarrow$ Based on the computation of $\mathbb{P}$(detection for $j$ visits)
    - Ecological effect of forest edge? Thus, there will be an edge effect on $\psi$

```{r real_life_data,eval=TRUE}
## Read data
## The data file "bluebug.txt" is available at
## http://www.vogelwarte.ch/de/projekte/publikationen/bpa/complete-code-and-data-files-of-the-book.html
data <- read.table("bluebug.txt", header = TRUE)

# Collect the data into suitable structures
y <- as.matrix(data[, 4:9])
y[y > 1] <- 1
edge <- data$forest_edge
dates <- as.matrix(data[, 10:15])
hours <- as.matrix(data[, 16:21])

# Standardize covariates
mean.date <- mean(dates, na.rm = TRUE)
sd.date <- sd(dates[!is.na(dates)])
DATES <- (dates-mean.date) / sd.date
DATES[is.na(DATES)] <- 0

mean.hour <- mean(hours, na.rm = TRUE)
sd.hour <- sd(hours[!is.na(hours)])
HOURS <- (hours-mean.hour) / sd.hour
HOURS[is.na(HOURS)] <- 0

last <- sapply(1:dim(y)[1],
               function(i) max(grep(FALSE, is.na(y[i, ]))))
#y[is.na(y)] <- 0

# Bundle data
bluebug.data <- list(y = y, R = nrow(y), T = ncol(y), edge = edge, DATES = DATES, HOURS = HOURS)

```

Objective: Fit the occupancy model and answer questions

Try to model 

* A linear effect of edge on $\psi$ (``edge`` is a binary covariate)

* A potentially quadratic effect of ``date`` and ``hour`` on $p$. These are the control variables, affect only the detection process, not the observation. 

**Note**

Remember the Poisson log-normal models for bird counts (or Normal for log-counts)? There we had *both control and ecological variables* affecting the same expected log-count $\lambda_{ijk}$. Here, the fact that there is a *latent variable* $Z_i$ (unobserved) at site $i$ means that we can differentiate processes that affect solely $p$ from processes that affect $\psi$ (for 0/1 data) or $\lambda$ (for counts). 

## Fitting the model

```{r fitting-the-model}

# Specify model in BUGS language
cat(file = "model.jags","
model {

# Priors
alpha.psi ~ ...
beta.psi ~ ...
alpha.p ~ ...
beta1.p ~ ...
beta2.p ~ ...
beta3.p ~ ...
beta4.p ~ ...

# Likelihood

# Ecological model for the partially observed true state
for (i in 1:R) {
 
   # Observation model for the observations
   for (j in 1:T) {
 
      } #j or t
   } #i

# Derived quantities
occ.fs <- ...                           # Number of occupied sites
mean.p <- ...                           # Sort of average detection
}
",fill = TRUE)

# Initial values
zst <- apply(y, 1, max, na.rm = TRUE)	# Good starting values crucial
inits <- function(){list(z = zst, alpha.psi=runif(1, -3, 3), alpha.p = runif(1, -3, 3))}

# Parameters monitored
params <- c("alpha.psi", "beta.psi", "mean.p", "occ.fs", "alpha.p", "beta1.p", "beta2.p", "beta3.p", "beta4.p")

# MCMC settings
ni <- 30000
nt <- 10
nb <- 20000
nc <- 3

# Call JAGS from R (BRT < 1 min)
out <- jags(bluebug.data, inits, params, "model.jags", n.chains = nc, n.thin = nt, n.iter = ni, n.burnin = nb, working.directory = getwd())

# Summarize posteriors
print(out, dig = 2)

# Posterior distribution of the number of occupied sites in actual sample
hist(out$BUGSoutput$sims.list$occ.fs, nclass = 30, col = "gray", main = "", xlab = "Number of occupied woodpiles (occ.fs)", xlim = c(9, 27))
abline(v = 10, lwd = 2) # The observed number
 
Pstar <- array(NA, dim = c(out$BUGSoutput$n.sims, 10))
x <- cbind(rep(1, 3000), rep(2, 3000), rep(3, 3000), rep(4, 3000), rep(5, 3000), rep(6, 3000), rep(7, 3000), rep(8, 3000), rep(9, 3000), rep(10, 3000)) 
for (i in 1:out$BUGSoutput$n.sims) {
   for (j in 1:10){
      Pstar[i,j] <- 1 - (1 - out$BUGSoutput$sims.list$mean.p[i])^j
      } #j
   } #i

boxplot(Pstar ~ x, col = "gray", las = 1, ylab = "Pstar", xlab = "Number of surveys", outline = FALSE)
abline(h = 0.95, lty = 2, lwd = 2)
 
par(mfrow = c(2, 1))
hist(plogis(out$BUGSoutput$sims.list$alpha.psi), nclass = 40, col = "gray", main = "Forest interior", xlab = "Occupancy probability", xlim = c(0, 1))
hist(plogis(out$BUGSoutput$sims.list$alpha.psi+ out$BUGSoutput$sims.list$beta.psi), nclass = 40, col = "gray", main = "Forest edge", xlab = "Occupancy probability", xlim = c(0, 1))
 
# Predict effect of time of day with uncertainty
mcmc.sample <- out$BUGSoutput$n.sims

original.date.pred <- seq(0, 60, length.out = 30)
original.hour.pred <- seq(180, 540, length.out = 30)
date.pred <- (original.date.pred - mean.date)/sd.date
hour.pred <- (original.hour.pred - mean.hour)/sd.hour
p.pred.date <- plogis(out$BUGSoutput$mean$alpha.p + out$BUGSoutput$mean$beta1.p * date.pred + out$BUGSoutput$mean$beta2.p * date.pred^2 )
p.pred.hour <- plogis(out$BUGSoutput$mean$alpha.p + out$BUGSoutput$mean$beta3.p * hour.pred + out$BUGSoutput$mean$beta4.p * hour.pred^2 )

array.p.pred.hour <- array.p.pred.date <- array(NA, dim = c(length(hour.pred), mcmc.sample))
for (i in 1:mcmc.sample){
   array.p.pred.date[,i] <- plogis(out$BUGSoutput$sims.list$alpha.p[i] + out$BUGSoutput$sims.list$beta1.p[i] * date.pred + out$BUGSoutput$sims.list$beta2.p[i] * date.pred^2)
   array.p.pred.hour[,i] <- plogis(out$BUGSoutput$sims.list$alpha.p[i] + out$BUGSoutput$sims.list$beta3.p[i] * hour.pred + out$BUGSoutput$sims.list$beta4.p[i] * hour.pred^2)
   }

# Plot for a subsample of MCMC draws
sub.set <- sort(sample(1:mcmc.sample, size = 200))

par(mfrow = c(2, 1))
plot(original.date.pred, p.pred.date, main = "", ylab = "Detection probability", xlab = "Date (1 = 1 July)", ylim = c(0, 1), type = "l", lwd = 3, frame.plot = FALSE)
for (i in sub.set){
   lines(original.date.pred, array.p.pred.date[,i], type = "l", lwd = 1, col = "gray")
   }
lines(original.date.pred, p.pred.date, type = "l", lwd = 3, col = "blue")

plot(original.hour.pred, p.pred.hour, main = "", ylab = "Detection probability", xlab = "Time of day (mins after noon)", ylim = c(0, 1), type = "l", lwd = 3, frame.plot = FALSE)
for (i in sub.set){
   lines(original.hour.pred, array.p.pred.hour[,i], type = "l", lwd = 1, col = "gray")
   }
lines(original.hour.pred, p.pred.hour, type = "l", lwd = 3, col = "blue")

```

```{r model-checks}
traceplot(out)
S<-ggs(as.mcmc(out)) #R2jags
S<-filter(S,Parameter != "deviance")
ggs_pairs(S)

```
