---
title: "Nonlinear models"
author: "F. Barraquand"
date: "January 09, 2023"
output:
  html_document:
  highlight: textmate
theme: united
toc: yes
toc_depth: 4
toc_float:
  collapsed: no
pdf_document:
  toc: yes
toc_depth: '4'
---
  
```{r setup, include=FALSE}
options(width = 300)
knitr::opts_chunk$set(cache = FALSE) 
library(R2jags)
library(knitr)
library(mcmcplots)
library(ggmcmc)

```

## Gompertz growth

```{r simulating data}
a = 15
b= 1.5
c = 0.15
timeindex=1:100
gompertz_growth = a*exp(-exp(b-c*timeindex)) + rnorm(100,0,1)
plot(timeindex,gompertz_growth)
data = list(T=100,y=gompertz_growth)
```

Objective: Fit in JAGS the Gompertz model corresponding to this simulated dataset. 


```{r fitting-the-model}
cat(file="gompertz.growth.txt","
model {
  # Priors
  a ~ dlnorm(1,0.01) #a>0
  b ~ dnorm(0,0.1)
  c ~ dnorm(0,0.1)
  sigma ~ dunif(0,10)
  tau<-pow(sigma,-2)
  
  # Likelihood
  for (t in 1:T){
    y[t] ~ dnorm(mu[t],tau)
    mu[t] <- a*exp(-exp(b-c*t))
  }

}    
    
")


# Inits function
inits <- function(){list(a = rlnorm(1, 0, 1),
                         b = rnorm(1,0,1), c = rnorm(1,0,1))}

# Parameters to estimate
params <- c("a","b","c","sigma")

# MCMC settings
nc <- 3  ;  ni <- 2000  ;  nb <- 1000  ;  nt <- 2

# Call JAGS, check convergence and summarize posteriors
out <- jags(data, inits, params, "gompertz.growth.txt", n.thin = nt, 
            n.chains = nc, n.burnin = nb, n.iter = ni)
print(out, dig = 3)     # Bayesian analysis


```

```{r model-checks}
traceplot(out)
S<-ggs(as.mcmc(out)) #R2jags
S<-filter(S,Parameter != "deviance")
ggs_pairs(S)

```

We check that in spite of the correlation we get nice curves

```{r checking-curves}
# We get the MCMC estimates for all iterations after burn-in
a_list<-out$BUGSoutput$sims.list$a
b_list<-out$BUGSoutput$sims.list$b
c_list<-out$BUGSoutput$sims.list$c
# We do a function that we want to represent
gompertz<-function(time,a,b,c){
  return(a*exp(-exp(b-c*time)))
  }
nlist = length(a_list)
nlist

Timemax<-100
library("scales")
plot(1:Timemax,gompertz(1:Timemax,a_list[1],b_list[1],c_list[1]),xlab="time",ylab="Length",col="blue",type="l")
lines(1:Timemax,gompertz(1:Timemax,a,b,c),col="black")
for (i in 2:nlist){
  lines(1:Timemax,gompertz(1:Timemax,a_list[i],b_list[i],c_list[i]),xlab="time",ylab="Length",col=alpha("blue",0.1))
}
```

Thus we see that here--and this is true for many nonlinear models--one should not be alarmed by the fact that two parameters are correlated: some can be correlated and yet the overall shape the curve is nicely simulated. Always check. (of course, correlation between parameters in the posteriors plots still means that parameters cannot be interpreted fully independently, especially if the precision in the posteriors is small.)
