---
title: "Analysis of Swiss Crested Tits population dynamics I"
author: "F. Barraquand - adapted from Kéry and Royle's Applied hierarchical modeling in Ecology"
date: "November 21, 2023"
output:
  html_document:
    highlight: textmate
    theme: united
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: '4'
---

```{r setup, include=FALSE}
options(width = 300)
knitr::opts_chunk$set(cache = FALSE) 
library(R2jags)
library(knitr)
library(mcmcplots)
```

```{r details-about-code, include=FALSE,eval=FALSE}
#  --------------- Original source: ---------------------------------------  
#   Applied hierarchical modeling in ecology
#   Modeling distribution, abundance and species richness using R and BUGS
#   Volume 2: Dynamic and Advanced models
#   Marc Kéry & J. Andy Royle
# Chapter 1 : RELATIVE ABUNDANCE MODELS FOR POPULATION DYNAMICS
# =============================================================
# Code from proofs dated 2020-08-18
# see https://github.com/mikemeredith/AHM_code
#  ---------------modified by FB starting 2021-05-23 ----------------------  
``` 

# Linear (mixed) models

## The model system and biological questions

We will be studying the [Swiss Crested Tit](https://en.wikipedia.org/wiki/European_crested_tit#/media/File:Lophophanes_cristatus_-_01.jpg). Its populations have been surveyed by birdwatchers at >250 sites, two to three times a year, from 1999 to 2016. Let's load the data and plot it: 

```{r load-and-plot-data}
dat<-read.csv("crestedTit.csv")
#str(dat)
dat$X<-NULL
# head(dat)
## Note: date991 is the date for year 99 and 1st survey, date992 second survey etc. 
C <- as.matrix(dat[,6:23]) # grab counts 1999:2016
#head(C)
year <- 1999:2016

# ~~~~ code to plot figure 1.3 ~~~~~~~~~~
matplot(year, t(C), main = "", type = "l", lty = 1, xlab = "Year",
    ylab = "Territory count (# individuals recorded)", lwd = 3, cex.lab = 1.5, frame = FALSE)

```


Our questions are the following:

* Is the variation in numbers greater per year or per site? 
* Does the species population dynamics exhibit a trend? Does the trend vary among sites?
* What are the effects of covariates like forest cover and elevation? (we know crested tits like trees)

Let's finish the data wrangling

```{r data-wrangling}
# Grab data for survey dates (for 2-3 surveys per year) and for duration
# Put into 3D array first, then summarize over reps within a year
nsite <- nrow(C)
nyear <- length(year)
datetmp <- as.matrix(dat[,24:77])
datefull <- array(datetmp, dim = c(nsite, 3, nyear))
durtmp <- as.matrix(dat[,78:131])
durfull <- array(durtmp, dim = c(nsite, 3, nyear))

# Get mean date of survey and mean survey duration for each site and year
date <- apply(datefull, c(1,3), mean, na.rm = TRUE)
dur <- apply(durfull, c(1,3), mean, na.rm = TRUE)
date[date == 'NaN'] <- NA
dur[dur == 'NaN'] <- NA
```

## Variation partitioning -- linear model on a transformed scale

The model that we want to fit to data is of the form 

$$ Y_{itk} = \mu + A_i + B_t +\epsilon_{itk}, \; \epsilon_{itk} \sim \mathcal{N}(0,\sigma^2)$$
with $A_i \sim \mathcal{N}(0,\sigma_{\text{site}}^2)$ and $B_t ~ \sim \mathcal{N}(0,\sigma_{\text{year}}^2)$. It is a *random effects* model, since all terms but the intercept are random variables.  

What does the data look like? Can we try to model this data using a transformation? 

```{r data-distribution}
hist(C) #highly skewed
hist(sqrt(C))
hist(log(C+1)) ## not perfect, but perhaps doable -> look at residuals. 
LC = log(C+1)
```

Let us select a subset of sites among the 250. Typically the ones with the most data (although this may affect some estimates, we'll come back to that later). 

```{r selecting sites}
# Check number of zero years in all sites
nzero <- apply(C, 1, function(x) sum(x == 0, na.rm = TRUE))
plot(sort(nzero)) # Make a graph of number of zero years
sum(nzero <= 1)   # 97 sites with at most 1 zero year
table(nzero)

# Bundle data with restriction on sites
sel <- nzero <= 1 # Select sites with <= 1 zero count
newM <- sum(sel)  # Define new number of sites
str(bdata <- list(C = C[sel,], M = newM, T = ncol(C[sel,])))
# List of 3
# $ C: int [1:97, 1:18] 3 5 9 13 12 2 3 3 6 10 ...
# $ M: int 97
# $ T: int 18

LC = log(C+1)
hist(LC[sel,]) 
str(bdatal <- list(LC = LC[sel,], M = newM, T = ncol(LC[sel,])))

```

This should make the model easier to fit and faster


```{r jags-code-LMM}
# Specify model in BUGS language
cat(file = "model_lmm.txt","
    model {

    # Priors
    mu ~ dnorm(0, 0.001) # Intercept
    for(i in 1:M){
      site[i] ~ dnorm(0,tau.site) # Random site effects
      }
    tau.site <- pow(sd.site, -2)
    sd.site ~ dunif(0,6) # dexp(1) other possibility
    for(t in 1:T){
      year[t] ~ dnorm(0,tau.year) # Random year effects
      }
    tau.year <- pow(sd.year, -2)
    sd.year ~ dunif(0,6)
    tau <- pow(sd, -2)
    sd  ~ dunif(0,6)

    # Likelihood
    for (i in 1:M){
      for(t in 1:T){
        LC[i,t] ~ dnorm(lambda[i,t],tau) # some distribution
        lambda[i,t] <- mu + site[i] + year[t] # some effects 
        resid[i,t]<-LC[i,t]-lambda[i,t]
        }
      }
    # Derived quantities
    for(t in 1:T){
      popindex[t] <- sum(lambda[,t])
      }
    }
    ")

# Initial values
inits <- function() list(mu = rnorm(1), site = rnorm(newM), year = rnorm(length(year)))

# Parameters monitored
params <- c("mu", "sd.site", "sd.year", "sd", "site", "year", "popindex","resid","lambda")

# MCMC settings
ni <- 10000; nt <- 10 ; nb <- 5000 ; nc <- 3 # core number for parallel computation

#to see progress do not use parallel computation (slower)
out_lmm <- jags(bdatal, inits, params, "model_lmm.txt", n.chains = nc,n.thin = nt, n.iter = ni, n.burnin = nb)

# A useful command is 
# print(out_lmm, dig=3) 
# however here we keep track of too many parameters (e.g. residuals,fitted values) thus will will avoid
out_lmm$BUGSoutput$mean$sd^2
out_lmm$BUGSoutput$mean$sd.site^2
out_lmm$BUGSoutput$mean$sd.year^2
total = out_lmm$BUGSoutput$mean$sd^2+ out_lmm$BUGSoutput$mean$sd.site^2+out_lmm$BUGSoutput$mean$sd.year^2
out_lmm$BUGSoutput$mean$sd^2/total
out_lmm$BUGSoutput$mean$sd.site^2/total
out_lmm$BUGSoutput$mean$sd.year^2/total

## Nota bene: possible to wire in such decomposition into the estimation, e.g. https://stateofther.github.io/post/rstan/WorkingWithStan_part2.html

```

So we found equal (mean) variance between sites and residuals, and less variance between years

```{r trace-and-density-plots,eval=F}

# traceplot() in native R2jags and other functions in library(coda)
traceplot(out_lmm, varname = "sd")
traceplot(out_lmm, varname = "popindex")

#also possible to do 
#traplot(out_lmm, parms = "sd") #but does not include sd.year, sd.site
#traplot(out_lmm, parms = "popindex")

denplot(out_lmm, parms = c("sd","sd.year","sd.site"))
denplot(out_lmm, parms = "popindex")
# other possible function
# densplot() in library(coda)
# densityplot()
```

Now we check the residuals

```{r plotting-residuals}
plot(out_lmm$BUGSoutput$mean$lambda,out_lmm$BUGSoutput$mean$resid, xlab="Fitted values", ylab="Residuals")
```

Note: do the same thing with lme4 ? Probably useful. In such a simple case there should not be any difference. 

### Exploring the trend - first with the random effects model

```{r popindex}
# Plot population size index
plot(year, out_lmm$BUGSoutput$mean$popindex, xlab = 'Year', ylab = 'Population index',
     pch = 16, frame = FALSE, type= 'b') #ylim = c(500, 1000), 
q02.5 = apply(out_lmm$BUGSoutput$sims.list$popindex,2,function(x){quantile(x,0.025)})
q97.5 = apply(out_lmm$BUGSoutput$sims.list$popindex,2,function(x){quantile(x,0.975)})
segments(year,q02.5, year, q97.5) #NB similar plot as before
```

Note: We could also have a simple LM model with *fixed* (non-random) effects: would not change much the estimation of the trend, but would not be able to estimate the variances, and some site effects might be poorly estimated. 

## Moving on to a fixed and random effect model

Here we want to include a general trend? Do we keep the year effect? 

[Todo by students]

```{r mixed-model}
str(bdatal <- list(LC = LC[sel,], M = newM, yr = year - mean(year), T = ncol(LC[sel,])))

cat(file = "model_lmm2.txt","
    model {

    # Priors
    mu ~ dnorm(0, 0.001) # Intercept
    for(i in 1:M){
      site[i] ~ dnorm(0,tau.site)  # Random site effects
      }
    tau.site <- pow(sd.site, -2)
    sd.site ~ dunif(0, 5)
    for(t in 1:T){
      year[t] ~ dnorm(0,tau.year) # Random year effects
      }
    tau.year <- pow(sd.year, -2)
    sd.year ~ dunif(0,3)
    tau <- pow(sd, -2)
    sd ~ dunif(0,3)
    
    # Prior on the slope
    gamma ~ dnorm(0,0.1)

    # Likelihood
    for (i in 1:M){
      for(t in 1:T){
        LC[i,t] ~ dnorm(lambda[i,t],tau)
        lambda[i,t] <- mu + site[i] + year[t] + gamma * yr[t] # mixed model
        }
      }
    # Derived quantities
    for(t in 1:T){
      popindex[t] <- sum(lambda[,t])
      }
    }
    ")

# Initial values
inits <- function() list(mu = rnorm(1), site = rnorm(newM), year = rnorm(length(year)),gamma=rnorm(1))

# Parameters monitored
params <- c("mu", "sd.site", "sd.year", "sd", "gamma", "popindex")

# MCMC settings
ni <- 10000; nt <- 10 ; nb <- 5000 ; nc <- 3 

out_lmm2 <- jags(bdatal, inits, params, "model_lmm2.txt", n.chains = nc,n.thin = nt, n.iter = ni, n.burnin = nb)
head(print(out_lmm2))

traceplot(out_lmm2, varname = "sd")
traceplot(out_lmm2, varname = "gamma")

denplot(out_lmm2, parms = c("sd","sd.year","sd.site"))
denplot(out_lmm2, parms = "gamma")

```

Now what is the actual average effect of that trend?

```{r effect-of-gamma}
yr = year - mean(year)
exp(out_lmm2$BUGSoutput$mean$mu+out_lmm2$BUGSoutput$mean$gamma*yr[1])-1 #average count at the start
exp(out_lmm2$BUGSoutput$mean$mu+out_lmm2$BUGSoutput$mean$gamma*yr[18])-1 #average count at the end

```


We now remove the random year effect

```{r mixed-model-bis}

cat(file = "model_lmm3.txt","
    model {

    # Priors
    mu ~ dnorm(0, 0.001) # Intercept
    for(i in 1:M){
      site[i] ~ dnorm(0,tau.site)  # Random site effects
      }
    tau.site <- pow(sd.site, -2)
    sd.site ~ dunif(0, 5)
 
    tau <- pow(sd, -2)
    sd ~ dunif(0,3)
    
    # Prior on the slope
    gamma ~ dnorm(0,0.1)

    # Likelihood
    for (i in 1:M){
      for(t in 1:T){
        LC[i,t] ~ dnorm(lambda[i,t],tau)
        lambda[i,t] <- mu + site[i] + gamma * yr[t] # mixed model
        }
      }
    # Derived quantities
    for(t in 1:T){
      popindex[t] <- sum(lambda[,t])
      }
    }
    ")


# Initial values
inits <- function() list(mu = rnorm(1), site = rnorm(newM),gamma=rnorm(1))

# Parameters monitored
params <- c("sd","sd.site","mu","gamma","popindex")

# MCMC settings
ni <- 10000; nt <- 10 ; nb <- 5000 ; nc <- 3 

out_lmm3 <- jags(bdatal, inits, params, "model_lmm3.txt", n.chains = nc,n.thin = nt, n.iter = ni, n.burnin = nb)


traceplot(out_lmm3, varname = "sd")
traceplot(out_lmm3, varname = "gamma")

denplot(out_lmm3, parms = c("sd","sd.year","sd.site"))
denplot(out_lmm3, parms = "gamma")

```

```{r comparison-gamma-estimates}
out_lmm2$BUGSoutput$mean$gamma
out_lmm3$BUGSoutput$mean$gamma
q02.5_lmm2 = quantile(out_lmm2$BUGSoutput$sims.list$gamma,0.025)
q97.5_lmm2 = quantile(out_lmm2$BUGSoutput$sims.list$gamma,0.975)
q02.5_lmm3 = quantile(out_lmm3$BUGSoutput$sims.list$gamma,0.025)
q97.5_lmm3 = quantile(out_lmm3$BUGSoutput$sims.list$gamma,0.975)
c(q02.5_lmm2,q97.5_lmm2)  
c(q02.5_lmm3,q97.5_lmm3)  
## This is something handy from package MCMCplot
#MCMCplot(out_lmm3,out_lmm2,params="gamma",ISB = TRUE, exact = FALSE)

```

Compare the predictions of models. 

```{r}
q02.5_lmm2 = apply(out_lmm2$BUGSoutput$sims.list$popindex,2,function(x){quantile(x,0.025)})
q97.5_lmm2 = apply(out_lmm2$BUGSoutput$sims.list$popindex,2,function(x){quantile(x,0.975)})
q02.5_lmm3 = apply(out_lmm3$BUGSoutput$sims.list$popindex,2,function(x){quantile(x,0.025)})
q97.5_lmm3 = apply(out_lmm3$BUGSoutput$sims.list$popindex,2,function(x){quantile(x,0.975)})

# Plot population index under models 1 and 2
plot(year-0.1, out_lmm2$BUGSoutput$mean$popindex, cex = 2, pch = 16, xlab = 'Year',
    ylab = 'Population index', col = 'red', type = 'b', frame = FALSE,
    main = 'red: with-year-effects, blue:without ') #ylim = c(500, 1000), 
segments(year-0.1, q02.5_lmm2, year-0.1, q97.5_lmm2, col = 'red')
points(year+0.1, out_lmm3$BUGSoutput$mean$popindex, cex = 2, pch = 16, col = 'blue', type = 'b')
segments(year+0.1, q02.5_lmm3, year+0.1, q97.5_lmm3, col = 'blue') #mind that the index is on the log-transformed scale

```

