---
title: "Analyses of piracy (kleptoparasitism) in salmon-eating Bald eagles"
author: "F. Barraquand - adapted from Mc Elreath's Statistical Rethinking"
date: "December 12, 2023"
output:
  html_document:
    highlight: textmate
    theme: united
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: '4'
---

```{r setup, include=FALSE}
options(width = 300)
knitr::opts_chunk$set(cache = FALSE) 
library(R2jags)
library(knitr)
library(mcmcplots)
library(MASS)
```

# Binomial GLM

The course part does logistic regression. Now we do something akin to an ANOVA, but with binary data. 


```{r eagles-exploring data}
data(eagles)
eagles
#P // Size of pirating eagle (L = large, S = small).
#A // Age of pirating eagle (I = immature, A = adult).
#V // Size of victim eagle (L = large, S = small).
 
eagles.data = list(N=nrow(eagles),y=eagles$y,z=eagles$n,
               P=as.numeric(eagles$P)-1,A=as.numeric(eagles$A)-1,
               V=as.numeric(eagles$V)-1)

```


We're in a similar situation to the turtles sex, but with categorical explanatory variables

```{r fitting the first model}

cat(file = "eagles.glm.txt","
model {

}
")

```

Fitting the model

```{r fitting-glm}
# Initial values

# Parameters to estimate

# MCMC settings
nc <- 3  ;  ni <- 2000  ;  nb <- 1000  ;  nt <- 2
# Call JAGS, check convergence and summarize posteriors
out <- jags(eagles.data, inits, params, "eagles.glm.txt",
            n.thin = nt, n.chains = nc, n.burnin = nb, n.iter = ni)
print(out, dig = 3)     # Bayesian analysis
```

Model checking

```{r checking}
traplot(out)
```


## Interpreting the model

We see the following effects: 
...

## A model with hierarchical priors (hyperprior)

Next put a hyperprior on $\beta$s. What are we transforming this model into? 

```{r new-model}

```

